Synthesis of Linear Functional Programs

### 2021-07-23 article/article.tex
Todos, tweaks, page composition :)

\mypara{Contributions and Outline} Despite their long-known potential
\cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018}
and strong proof-theoretic foundations
\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
synthesis with linear types combined with other advanced typing
features has generally been overlooked in the literature.  In this
work we present a framework for synthesis with linear types extended
with recursive algebraic data types, parametric polymorphism and
refinements.
% TODO: extensible? proving feasability of... ? showing potential for ... ?
We first introduce linear types as specifications
and outline the synthesis process, leveraging linearity, by example
(\S~\ref{sec:overview}). We then present the formal system driving the
synthesis (\S~\ref{sec:formal_system}) and describe the architecture
of our framework named \synname, examining technical details and key
implementation challenges (\S~\ref{sec:architecture}). Finally, 
we evaluate our work through expressiveness and performance benchmarks
(\S~\ref{sec:evaluation}) and discuss related work
(\S~\ref{sec:related}). Appendix~\ref{sec:background} covers background
concepts such as \emph{linear logic} and \emph{sequent calculus}.
Appendix~\ref{sec:final_system} presents the final set of inference
rules, putting together everything seen in \S~\ref{sec:formal_system}.
Appendix~\ref{sec:examples} lists concrete examples of synthesis with \synname.

@@

Linear types make for more precise specifications than simple types
because information on which resources must necessarily be used is
encoded in the type. For instance the type $\mathsf{Int} \lolli
\mathsf{Int} \lolli \mathsf{Int}$ denotes a function that \emph{must}
use its two integer arguments to produce an integer.
Their preciseness also affects the search space:
all programs where a linear resource is used non-linearly (i.e. not
\emph{exactly once}) are ill-typed. With linearity built into the
synthesis process, usage of a linear proposition more than once is not
considered, and unused propositions are identified during synthesis,
constraining the space of valid programs.  % TODO: Reword?

@@

\mypara{ADTs} We reiterate the expressiveness of linear types: by requiring
some types to be used linearly (consumed), we can assure, e.g., the
deconstruction of an ADT. As such, the specification that identifies a function
that takes as arguments an unrestricted ($\bang$) linear function\footnote{might be used non-linearly, but its parameter is used
linearly in its body} 
that converts $a$s in $b$s, a list of $a$s, and produces a
list of $b$s, written $\bang(a \lolli b) \lolli List\ a \lolli List\ b$, isn't
satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}. To synthesize
a valid program from it, \mintinline{haskell}{(List a)} has to be deconstructed,
and its constructor arguments used. \synname\ would output the following
function (\emph{map}) without relying on any additional information:
%
% TODO: Aqui não estou bem a falar de polymorphism mas parece estranho não por
% visto que falo dos outros temas? se calhar faço um à parte?
%
\begin{minted}{haskell}
map c d = let !e = c in
  case d of
      Nil -> Nil
    | Cons f -> let g*h = f in Cons (e g, map (!e) h);
\end{minted}

\mypara{Refinements} Refinements are logical predicates
which must hold for any element that inhabits a given type. The \synname\
language supports simple refinements on integers with arithmetic
expressions in the predicates. They assert the robustness of our main
framework, i.e., how it is amenable to significant additions without
necessarily interfering with the overall approach.
%
As an example, from the specification $x \{Int\} \lolli y \{Int\} \lolli z
\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$ (a function that takes three
integers and produces an integer that satisfies the given predicate), the
program\ \mintinline{haskell}{(\a -> \b -> \c -> ((0 - 1) * a) + (c * b))} is
synthesized.

@@

Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
more intricate example of synthesis: given the linear type signatures for array
primitives (\emph{newMArray} passes a new mutable array to a function that uses
it linearly, \emph{write} takes a mutable array and writes a pair to it,
\emph{freeze} consumes a mutable array and produces an imutable array,
\emph{foldl} has the default definition) we synthesize a function
``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
follows:
%
\begin{minted}{haskell}
foldl :: (a -o b -o a) -> a -o (List b) -o a;
newMArray :: Int -> (MArray a -o !b) -o b;
write :: MArray a -o (Int * a) -> MArray a;
freeze :: MArray a -o !(Array a);
synth array :: Int -> List (!(Int * a)) -> Array a
    | using (foldl) | depth 3;
\end{minted}
%
Matching exactly the linear definition for
\emph{array} from~\cite{Bernardy_2018}, \synname\
outputs (in a fraction of a second):
%
\begin{minted}{haskell}
array b c = let !d = b in let !e = c in
    newMArray (!d) (\j -> freeze (foldl (!write) j e));
\end{minted}

@@

%\todo[inline]{Aqui parece-me que deves comecar por explicar o que vais fazer
%nesta seccao. Sinto falta da sintaxe da linguagem, possivelmente
%incrementalmente. Comecava por listar os tipos/termos do lambda linear
%``simples'', depois em cada subseccao acrescentava as construcoes novas e como a
%sintese funciona para esse contexto. Enfatizar que a novelty é a combinacao
%destas features todas.}

%The development of \synname\ was based on an almost direct reading of concrete
%inference rules and theoretical principles (~\cite{frank pfenning notes});
%%
%\todo{aqui queres dizer que começaste por te basear num sistema de focusing para
%a logica linear (era sequer linear a formulacao que comecaste a seguir?) O que
%são theoretical principles?}
%%
%and work on its environment was also guided by formal systems and methods
%(e.g.~Hindley-Milner type inference~\cite{}).
%%
%\todo{esta frase parece-me desnecessaria. O que queres dizer neste paragrafo,
%parece-me, é que partes de um sistema focused da literatura que depois vais
%extender, inspirando te em tecnicas standard de proof search / inferencia.}

%However, with the addition of recursion and richer types to the synthesis and
%language, formal guidelines
%%
%\todo{nao sao formal guidelines, nao existem trabalhos a combinar estas features
%todas, essa parte é importante. Além de que ao incluir estas coisas deixamos de
%estar completamente em proof search, pois muitas destas coisas (e.g. recursao)
%nao sao necessariamente sound logicamente.  Tenta usar termos especificos.
%Richer types? Richer como?}
%%
%to support the implementation become scarcer, or more complex.
%%
%\todo{nao existem ou sao complicados? ambos? citacao?}
%%
%Accordingly, rather than derived
%from rules, solutions appeared first as techniques to tame the extensions'
%inherent complex problems (such as infinite recursion), and we now present
%those techniques as rules.
%%
% \todo{ Nao percebo o que queres dizer aqui. O que sao explicit rules e implicit rules?
% Ou há regras ou nao há. O que estas a
% dizer é que inventaste tecnicas de sintese para estes casos, que correspondem
% as regras que vais apresentar. E tao explicito ou implicito como o que esta na
% literatura.}

@@

In this section we formalize  the techniques that guide
synthesis from our more expressive specifications, alongside the already well
defined rules that model the core of the synthesizer, putting together a
\emph{sound} set of inference rules that characterizes our framework
for linear synthesis of
recursive programs from specifications with the select richer types,
% TODO: Better?....
and describes the system in enough detail for the synthesizer to be reproducible
by a theory-driven implementation.

We note that a \emph{sound} set of rules guarantees we cannot synthesize
incorrect programs; and that the valid programs derivable through them
reflect the subjective trade-offs we committed to. Different
choices and approaches outside the core might lead to completely
distinct synths and spaces of valid programs.

@@

The core language is a simply-typed linear $\lambda$-calculus with linear
functions ($\lolli$), additive ($\with$) and multiplicative
($\tensor$) pairs, \todo{qual a diferenca}
multiplicative unit (\textbf{1}), additive sums ($\oplus$)
and the exponential modality ($\bang$) (to internalize unrestricted
use of variables). The syntax of terms $(M,N)$ and types $\tau, \sigma$ is given below:

%
% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
% usá-los como tipos gerais? NAO :)

\[
  \begin{array}{lclcc}
    M,N & ::= & u, v & \quad & \\
        & \mid & \lambda x.M \mid M\, N & & (\lolli) \\
        & \mid & M \with N\ \vert\ \mathsf{fst}\ M\ \vert\
                 \mathsf{snd}\ M && (\with) \\
        & \mid & M \tensor N\ \vert\ \llet{u\tensor v = M}{N} && (\tensor)\\
        & \mid & \star\ \vert\ \llet{\star = M}{N} && (\one) \\
        & \mid & \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
                 N_2) && (\oplus)\\
        & \mid & \bang M\ \vert\ \llet{\bang u = M}{N} && (\bang)\\[0,5em]
    \tau, \sigma & ::= & a\ \vert\ \tau \lolli \sigma\ \vert\ \tau \with \sigma\
    \vert\ \tau \tensor \sigma\ \vert\ \mathbf{1}\ \vert\ \tau \oplus \sigma\ \vert\ \bang \tau
  \end{array}
  \]

%\vspace{-0.5cm}
% \[
% \begin{tabular}{lclc}
% $M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
%     & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
%     & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
%     & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
%     & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
%     & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
%     M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
%     N_2)$ & $\oplus$ \\
%     & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
%     \\
% $\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
%     \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
% \end{tabular}
% \]

%
% TODO: Onde falamos de resource management? ( 
%
In intuitionistic sequent calculi, each connective has a so-called \emph{left}
and a \emph{right} rule, which effectively define how to decompose an ambient
assumption of a given proposition and how to prove a certain proposition is
true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible and
non-invertible inference rules. Andreoli observed that the
connectives of linear logic can be divided into two categories, dubbed
synchronous and asynchronous. Asynchronous connectives are those whose
right rules are \emph{invertible}, i.e. they can be applied eagerly
during proof search without losing provability and so the order in
which these rules are applied is irrelevant, and whose left rules
are not invertible. Synchronous connectives are dual. The asynchronous \todo{Escrever isto depois
de apresentar o linear $\lambda$-calculus?} 
connectives are $\lolli$ and $\with$ and the synchronous are
$\tensor,\textbf{1},\oplus,\bang$.

@@

% TODO: Explicar vários judgments?
% \todo{Primeiro explica os varios judgments (as ``formas'' das
%   conclusoes das regras) e o que significam as varias componentes,
%   nomeadamente a parte da gestao de recursos.  Depois podes explicar
%   regras concretas como fazes a seguir. Tambem seria util dizer quais
%   os tipos que sao invertiveis a esquerda/direita, pois o leitor nao
%   faz ideia do que estas a dizer.}
%
% Esta parte em geral está a ficar pouco coerente, com a introdução do focusing,
  % a suposta introdução dos judgments, e a introdução da gramática... deixa

  %
\todo[inline]{Context management -- To handle the context splitting
  required ... we augment the judgments above with .... where ....}

Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (%heavily
inspired
by~\cite{FPnotes...}) -- in which the
rule $\lolli$R is read: to synthesize a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthesize a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

We begin with the right invertible rules:\todo{que decompoem a direita}
\begin{mathpar}

@@

When we reach a non-invertible proposition on the right, we start inverting the
$\Omega$ context. The rule to transition to inversion on the left is:
\begin{mathpar}
    \infer*[right=($\Uparrow$R)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
    right asynchronous}}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
\end{mathpar}

Followed by the left invertible rules, which ...\todo{fazem o que?}:
\[
  \begin{array}{c}

@@

  The decision rules are followed by either left or right focus rules.
  To illustrate, consider the $\lolli$L left focus rule. The rule
  states that to produce a program of type $C$ while left focused on
  the function $x$ of type $A\lolli B$, we first check that we can produce a
  program of type $C$ by using $B$. If this succeeds in producing some
  program $M$, this means that we can apply $x$ to solve our
  goal. We now synthesize a program $N$ of type $A$, switching to the
  right inversion judgment ($\Uparrow$). To construct the overall
  program, we replace in $M$ all occurrences of variable $y$ with the
  application $x\,N$. The remaining left rules follow a similar
  pattern. \todo{The focus right rules \dots }

  \begin{mathpar}
      \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash
      M\{(\textrm{snd}\ x)/y\} : C}
    \and
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    { }
    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\[
  \begin{array}{c}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
    \qquad
     \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}\\[0.5em]
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}

@@

\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
is a tagged sum of any type, i.e. a named type that can be instantiated by one of many
tags (or constructors) that take some value of  a fixed type, which might
be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
for constructors with an arbitrary number of parameters. In the \synname\
language, the programmer can define custom ADTs; as an example, we show the
definition of an ADT which holds zero, one, or two
values of type $A$, using the syntax: \mintinline{haskell}{data Container
= None 1 | One A | Two (A * A)}. The grammar is extended as (where $C$
is and $T$ is \dots):\todo{look here}
\[
\begin{tabular}{lclc}
    $M,N$ & $\ ::=\ $ & $\dots\ \vert\ \emph{C}_n\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of}\ \dots\ \vert\ \emph{C}_n\ u \Rightarrow N)$ \\
    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ T$ \\
 %    $ADT$ & $\ ::=\ $ & (\textsf{data} $\langle\emph{name}\rangle$\ $=$\ $\emph{Cons}_1\ \tau$\
% $\vert$\ $\dots$\ $\vert$\ $\emph{Cons}_n\ \tau$) \\
\end{tabular}
\]
%TODO: Como definir gramática dos ADTs?

The semantics of ADTs relate to those of the plus ($\oplus$) type
-- both are additive disjunctions.  To construct a value of an ADT we
must use one of its constructors, similar to the way $\oplus$ requires
only proof of either the left or right type it consists
of. Analogously, we can deconstruct a value of an ADT via pattern
matching on its constructors, where all branches of the pattern match
must have the same type -- akin to the left rule for the $\oplus$
connective. In effect, the inference rules for a simple ADT are a
generalized form of the $\oplus$ rules.  Therefore, there's one left
rule for ADTs, and an arbitrary number of right rules, one for each
constructor:

>>>>

\mypara{Contributions and Outline} Despite their long-known potential
\cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018} and
strong proof-theoretic foundations
\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
synthesis with linear types combined with other advanced typing features has
generally been overlooked in the literature.  In this work we present a
framework for synthesis with linear types extended with recursive algebraic data
types, parametric polymorphism and refinements.
% TODO: extensible? proving feasability of... ? showing potential for ... ?
We first introduce linear types as specifications and outline the synthesis
process, leveraging linearity, by example (\S~\ref{sec:overview}). We then
discuss the formal system driving the synthesis (\S~\ref{sec:formal_system}) and
describe the architecture of our framework named \synname, examining technical
details and key implementation challenges (\S~\ref{sec:architecture}). Finally,
we evaluate our work through expressiveness and performance benchmarks
(\S~\ref{sec:evaluation}) and discuss related work (\S~\ref{sec:related}).
Appendix~\ref{sec:background} covers background concepts such as \emph{linear
logic} and \emph{sequent calculus}. Appendix~\ref{sec:final_system} presents
the final set of inference rules. Appendix~\ref{sec:examples} lists concrete examples
of synthesis with \synname.

@@

Linear types make for more precise specifications than simple types
because information on which resources must necessarily be used is
encoded in the type. For instance the type $\mathsf{Int} \lolli
\mathsf{Int} \lolli \mathsf{Int}$ denotes a function that \emph{must}
use its two integer arguments to produce an integer.
Their preciseness also affects the search space:
all programs where a linear resource is used non-linearly (i.e. not
\emph{exactly once}) are ill-typed. With linearity built into the
synthesis process, usage of a linear proposition more than once is not
considered, and unused propositions are identified during synthesis,
constraining the space of valid programs.

@@

\mypara{ADTs} We reiterate the expressiveness of linear types: by requiring
some types to be used linearly (consumed), we can assure, e.g., the
deconstruction of an ADT. As such, the specification that identifies a function
that takes as arguments an unrestricted ($\bang$) linear function\footnote{might be used non-linearly, but its parameter is used
linearly in its body} 
that converts $a$s in $b$s, a list of $a$s, and produces a
list of $b$s, written $\bang(a \lolli b) \lolli List\ a \lolli List\ b$, isn't
satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}. To synthesize
a valid program from it, \mintinline{haskell}{(List a)} has to be deconstructed,
and its constructor arguments used. \synname\ would output the following
function (\emph{map}) without relying on any additional information:
%
% TODO: Aqui não estou bem a falar de polymorphism mas parece estranho não por
% visto que falo dos outros temas? se calhar faço um à parte?
%
\begin{minted}{haskell}
map !e d = case d of
      Nil -> Nil
    | Cons g*h -> Cons (e g, map (!e) h);
\end{minted}

\mypara{Refinements} Refinements are logical predicates
which must hold for any term that inhabits a given type. The \synname\
language supports simple refinements on integers with arithmetic
expressions in the predicates. They assert the robustness of our main
framework, i.e., how it is amenable to significant additions without
necessarily interfering with the overall approach.
%
As an example, from the specification $x \{Int\} \lolli y \{Int\} \lolli z
\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$ (a function that takes three
integers and produces an integer that satisfies the given predicate), the
program\ \mintinline{haskell}{(\a -> \b -> \c -> ((0 - 1) * a) + (c * b))} is
synthesized.

@@

Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
more intricate example of synthesis: given the linear type signatures for array
primitives (\emph{newMArray} passes a new mutable array to a function that uses
it linearly, \emph{write} takes a mutable array and writes a pair to it,
\emph{freeze} consumes a mutable array and produces an imutable array,
\emph{foldl} has the default definition) we synthesize a function
``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
follows:
%
\begin{minted}{haskell}
foldl :: (a -o b -o a) -> a -o (List b) -o a;
newMArray :: Int -> (MArray a -o !b) -o b;
write :: MArray a -o (Int * a) -> MArray a;
freeze :: MArray a -o !(Array a);
synth array :: Int -> List (!(Int * a)) -> Array a
    | using (foldl) | depth 3;
\end{minted}
%
Matching the linear definition for \emph{array} from~\cite{Bernardy_2018}, \synname\
outputs (\emph{in 0.1s}):
%
% TODO: Make it actually output this... need pattern matching on lambdas + minimization...
\begin{minted}{haskell}
array !d !e = newMArray (!d) (\j -> freeze (foldl (!write) j e));
\end{minted}

@@

@@

In this section we formalize  the techniques that guide
synthesis from our more expressive specifications, alongside the already well
defined rules that model the core of the synthesizer, putting together a
\emph{sound} set of inference rules that characterizes our framework
for linear synthesis of
recursive programs from specifications with the select richer types,
and describes the system in enough detail for the synthesizer to be reproducible
by a theory-driven implementation.
%
We note that a \emph{sound} set of rules guarantees we cannot synthesize
incorrect programs; and that the valid programs derivable through them
reflect the subjective trade-offs we committed to. Different
choices and approaches outside the core might lead to completely
distinct synths and spaces of valid programs.

@@

The core language is a simply-typed linear $\lambda$-calculus with linear
functions ($\lolli$), additive ($\with$) and multiplicative
($\tensor$) pairs
(denoting alternative \emph{vs} simultaneous occurrence of resources),
multiplicative unit (\textbf{1}), additive sums ($\oplus$)
and the exponential modality ($\bang$) (to internalize unrestricted
use of variables). The syntax of terms $(M,N)$ and types $(\tau, \sigma)$ is given below:
\[
  \begin{array}{lclcc}
    M,N & ::= & u, v & \quad & \\
        & \mid & \lambda x.M \mid M\, N & & (\lolli) \\
        & \mid & M \with N\ \vert\ \mathsf{fst}\ M\ \vert\
                 \mathsf{snd}\ M && (\with) \\
        & \mid & M \tensor N\ \vert\ \llet{u\tensor v = M}{N} && (\tensor)\\
        & \mid & \star\ \vert\ \llet{\star = M}{N} && (\one) \\
        & \mid & \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
                 N_2) && (\oplus)\\
        & \mid & \bang M\ \vert\ \llet{\bang u = M}{N} && (\bang)\\[0,5em]
    \tau, \sigma & ::= & a\ \vert\ \tau \lolli \sigma\ \vert\ \tau \with \sigma\
    \vert\ \tau \tensor \sigma\ \vert\ \mathbf{1}\ \vert\ \tau \oplus \sigma\ \vert\ \bang \tau
  \end{array}
\]

In intuitionistic sequent calculi, each connective has a so-called \emph{left}
and a \emph{right} rule, which effectively define how to decompose an ambient
assumption of a given proposition and how to prove a certain proposition is
true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible and
non-invertible inference rules. Andreoli observed that the connectives of linear
logic can be divided into two categories, dubbed synchronous and asynchronous.
Asynchronous connectives are those whose right rules are \emph{invertible}, i.e.
they can be applied eagerly during proof search without losing provability and
so the order in which these rules are applied is irrelevant, and whose left
rules are not invertible. Synchronous connectives are dual. The asynchronous
connectives are $\lolli$ and $\with$ and the synchronous are
$\tensor,\textbf{1},\oplus,\bang$.

@@

To handle the context splitting required to prove subgoals, we augment the
judgments above using Hodas and Miller's resource management technique where a pair of
input/output linear contexts is used to propagate the yet unused linear
resources across subgoals; e.g. the left inversion judment is written
$\Gamma; \Delta/\Delta'; \Omega \Uparrow\ \nolinebreak\vdash A$ where $\Delta$ is the input
linear context and $\Delta'$ is the output one.

Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system
(inspired by~\cite{FPnotes...}) -- in which the
rule $\lolli$R is read: to synthesize a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthesize a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

We start with right invertible rules, which decompose the goal proposition until it's synchronous.
%
\begin{mathpar}

@@

When we reach a non-invertible proposition on the right, we start inverting the
$\Omega$ context. The rule to transition to inversion on the left is:
%
\begin{mathpar}
    \infer*[right=($\Uparrow$R)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
    right asynchronous}}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
\end{mathpar}

We follow with left invertible rules for asynchronous connectives, which
decompose asynchronous propositions in $\Omega$.
\[
  \begin{array}{c}

@@

The decision rules are followed by either left or right focus rules.
To illustrate, consider the $\lolli$L left focus rule. The rule
states that to produce a program of type $C$ while left focused on
the function $x$ of type $A\lolli B$, we first check that we can produce a
program of type $C$ by using $B$. If this succeeds in producing some
program $M$, this means that we can apply $x$ to solve our
goal. We now synthesize a program $N$ of type $A$, switching to the
right inversion judgment ($\Uparrow$). To construct the overall
program, we replace in $M$ all occurrences of variable $y$ with the
application $x\,N$. The remaining left rules follow a similar
pattern. The right focus rules are read similarly to right inversion ones,
albeit the goal and sub-goals are under focus (except for $\bang R$).
%
\begin{mathpar}
      \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash
      M\{(\textrm{snd}\ x)/y\} : C}
    \and
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    { }
    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\[
\begin{array}{c}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
    \qquad
     \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}\\[0.5em]
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}

@@

\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
is a tagged sum of any type, i.e. a named type that can be instantiated by one of many
tags (or constructors) that take some value of  a fixed type, which might
be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
for constructors with an arbitrary number of parameters. In the \synname\
language, the programmer can define custom ADTs; as an example, we show the
definition of an ADT which holds zero, one, or two
values of type $A$, using the syntax: \mintinline{haskell}{data Container
= None 1 | One A | Two (A * A)}. The grammar is extended as (where $C$ is an ADT
constructor and $T$ is an ADT):
\[
\begin{array}{lclc}
    M,N & ::= & \dots\ \vert\ \emph{C}_n\ M\ \vert\ (\mathsf{case}\ M\ \mathsf{of}\ \dots\ \vert\ \emph{C}_n\ u \Rightarrow N) \\
    \tau, \sigma & ::= & \dots\ \vert\ T \\
 %    $ADT$ & $\ ::=\ $ & (\textsf{data} $\langle\emph{name}\rangle$\ $=$\ $\emph{Cons}_1\ \tau$\
% $\vert$\ $\dots$\ $\vert$\ $\emph{Cons}_n\ \tau$) \\
\end{array}
\]
